---
title: "R Notebook"
output: 
  html_notebook
---

# 1. Preparation

## 1.1 Libraries

```{r}
library(tidyverse)
library(knitr)
```

## 1.2 Load data

```{r}
# US demographic data
data <- read_csv("acs2015_county_data.csv")

# US abbreviations
abbreviations <- read_tsv("US_state_abbreviations.txt")

# Join data
data <- full_join(data, abbreviations, by = "State")
```

This data is taken from the _2015 American Community Survey 5-year estimates_, 
the dataset can be found [here](https://www.kaggle.com/muonneutrino/us-census-demographic-data/home).



```{r }
# Structure
glimpse(data)
```


# 2. Analysis

# 2.1 Income

# 2.1.2 Income per state

```{r}
# Summarise population and income per state
state_data <- data %>%
  group_by(State) %>%
  summarise(Abbreviation = unique(Abbreviation),
            Total_pop = sum(TotalPop, na.rm = TRUE),
            Income_sd = sd(Income),
            Income = mean(Income, na.rm = TRUE),
            Income_per_cap = mean(IncomePerCap, na.rm = TRUE),
            Income_per_cap_sd = sd(IncomePerCap,na.rm = TRUE),
            Poverty = sum(Poverty / 100 * TotalPop, na.rm = TRUE),
            Poverty_percent = Poverty / Total_pop * 100,
            Unemployment = sum(Unemployment / 100 * TotalPop, na.rm = TRUE),
            Unemployment_percent = Unemployment / Total_pop * 100,
            White = sum(White / 100 * TotalPop, na.rm = TRUE),
            White_percent = White / Total_pop * 100,
            Black = sum(Black / 100 * TotalPop, na.rm = TRUE),
            Black_percent = Black / Total_pop * 100,
            Hispanic = sum(Hispanic / 100 * TotalPop, na.rm = TRUE),
            Hispanic_percent = Hispanic / Total_pop * 100,
            Asian = sum(Asian / 100 * TotalPop, na.rm = TRUE),
            Asian_percent = Asian / Total_pop * 100,
            Native = sum(Native / 100 * TotalPop, na.rm = TRUE),
            Native_percent = White / Total_pop * 100,
            Pacific = sum(Pacific / 100 * TotalPop, na.rm = TRUE),
            Pacific_percent = Pacific / Total_pop * 100)

# At least texas has a NA in Income

# Plot
plot_histogram(state_data, Income, 
               x_label = "Income ($)",
               title_string = "Histogram showing the median income for each state",
               subtitle_string = "Puerto Rico is an outlier")
plot_histogram(state_data, Income_per_cap,
               x_label = "Income per capita ($)",
               title_string = "Histogram showing the  income per capita for each state",
               subtitle_string = "Puerto Rico and Washington DC are outliers")

# Is income and income per capita correlated?
plot_scatterplot_text(state_data, Income, Income_per_cap, 
                      y_label = "Income per capita ($)", x_label = "Income ($)", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "Median income and income per capita show a clear correlation for US states",
                      subtitle_string = "Washington DC has a higher income per capita then predicted by the model")

# Is income and population correlatad?
plot_scatterplot_text(state_data, Total_pop, Income,  
                      y_label = "Income ($)", x_label = "Population", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "Median income and population have no correlation for US states")

# Is income and poverty correlated?
plot_scatterplot_text(state_data, Poverty_percent, Income,  
                      y_label = "Income ($)", x_label = "Percentage of population living in poverty", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "Median income and share of population have a negative correlation",
                      subtitle_string = "Puerto Rico is an outlier again")

# Is income and unemployment correlated?
plot_scatterplot_text(state_data, Unemployment_percent, Income,  
                      y_label = "Income ($)", x_label = "Percentage of population unemployed", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "Median income and unemployment correlates negatively",
                      subtitle_string = "Puerto Rico is poorer than any other US state")
```

## 2.2 Population

# 2.2.1 Population and electoral college votes

```{r}
# Read electoral college data
ec_data <- read_tsv("electoral_college.txt") %>%
  rename(ec_votes = `Number of Electoral Votes`)

# Join with state data
state_data <- full_join(state_data, ec_data, by = "State")

# How does state population and electoral college votes correlate?
plot_scatterplot_text(state_data, Total_pop, ec_votes,  
                      y_label = "Electoral votes", x_label = "Population", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "Median income and unemployment correlates negatively",
                      subtitle_string = "Puerto Rico is poorer than any other US state")
```

```{r}
# Print population table sorted to see where Puerto Rico is
kable(state_data %>%
  select(State, Total_pop, ec_votes) %>%
  mutate(rank = 53 -rank(Total_pop)) %>%
  select(State, Total_pop, ec_votes, rank) %>%
  arrange(desc(Total_pop)), format = "pandoc")
```

Puerto Rico is the 30th most populated state. I'm interested how many electoral
votes Puerto Rico would be granted if it became a state

```{r}
# Calculate and print simple ec votes according to poulation with Puerto Rico included
state_ec_data <- state_data %>%
  select(State, Abbreviation, Total_pop, ec_votes) %>%
  mutate(rank = 53 -rank(Total_pop),
         US_total_pop = sum(Total_pop),
         pop_percent = Total_pop / US_total_pop * 100,
         Total_ec_votes = sum(ec_votes, na.rm = TRUE),
         new_ec_votes = round(pop_percent / 100 * Total_ec_votes)) %>%
  select(State, Abbreviation, Total_pop, pop_percent, ec_votes, new_ec_votes, rank)

kable(state_ec_data %>%
        arrange(desc(Total_pop)), format = "pandoc")
```

This is obiously not exactly how this is done, this method would give more votes
to more popolous states and less to less popolous states.

After some reading I found out that each state gets 1 vote per senator (2 in 
total) and 1 per representatives in the congress. So if we subtract 104 (the 
number of senators if both Puerto Rico and Washington DC were considered to be 
states) from the total number of electoral votes and then distribute the 
remaining votes proportionally to their population.

```{r}
state_ec_data <- state_ec_data %>%
  mutate(total_ec_votes = sum(ec_votes, na.rm = TRUE) - 100,
         correct_ec_votes = 2 + round(pop_percent / 100 * total_ec_votes))

kable(state_ec_data %>%
        arrange(rank), format = "pandoc")
```

This looks about right! Let's calculate population per electoral vote:

```{r}
state_ec_data <- state_ec_data %>%
  mutate(population_per_vote = Total_pop / correct_ec_votes,
         population_per_vote_proportional = Total_pop / new_ec_votes )

kable(state_ec_data %>%
        arrange(desc(Total_pop)), format = "pandoc")
```

Let's see the correlation between number of votes and population.

```{r}
# Correct ec votes
plot_scatterplot_text(state_ec_data, Total_pop, population_per_vote,  
                      y_label = "Population per vote", x_label = "Population", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "Smaller state have more votes per capita than larger states")

# Proportional ec votes
plot_scatterplot_text(state_ec_data, Total_pop, population_per_vote_proportional,  
                      y_label = "Population per vote", x_label = "Population", 
                      text = Abbreviation, smooth = TRUE,
                      title_string = "A proportional electoral college system creates a large variance\nin smaller states",
                      subtitle_string = "This is probably due to the rounding of decimal numbers")
```

I'd say both these plots are evidence that a completely proportional system that
would not depend on electoral college votes would be the most fair system for
US presidential elections. But at least with the proportional electoral system
most states have about 600 000 inhabitants per vote.


