---
title: "US demographics new"
author: "Erik Ländström"
date: "25 March 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Read libraries

```{r library}
library(tidyverse)
library(knitr)
library(kableExtra)
```

# 2. Process and tidy data

## 2.1 Read data

Read demographics data from the _2015 American Community Survey 5-year estimates_, 
the dataset can be found [here](https://www.kaggle.com/muonneutrino/us-census-demographic-data/home). 

```{r read_data}
# Read data
demographics_data <- read_csv("acs2015_county_data.csv")

# Take a glimpse at the data
glimpse(demographics_data)
```

Read state abbreviations and join with county data.

```{r state_abbs}
# Read state abbreviations
state_abbs <- read_tsv("US_state_abbreviations.txt")

# Take a look at the state abbreviations
state_abbs %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

```{r join_county_abbs}
# Join demographics data with state abbs
demographics_data <- full_join(demographics_data, state_abbs, by = "State")
```

# 3. Data analysis

# 3.1 Population

```{r func1, echo = FALSE}
plot_histogram <- function(data, column, x_label,
                           title_string = "Histogram", subtitle_string = NULL,
                           caption_string = NULL) {
  
  # Libraries
  library(tidyverse)
  
  # quote column variable
  column <- enquo(column)
  
  # Create a vector of the data in the column
  vec <- as_vector(data %>%
                     select(!!column))
  
  # Calculates the optimal number of bins for one sample according to the Freedman-Diaconis rule
  b <- diff(range(vec, na.rm = TRUE)) / (2 * IQR(vec, na.rm = TRUE) / length(vec)^(1 / 3))
  
  # Plot histogram
  plot <- data %>%
    ggplot(aes(!!column)) +
    geom_histogram(bins = b, fill = "grey", color = "black") +
    xlab(x_label) +
    ylab("Count") +
    labs(
      title = title_string,
      subtitle = subtitle_string,
      caption = caption_string
    ) +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(color = "black")
    )
  
    plot +
      scale_y_continuous(expand = c(0,0),
                         limits=c(0,max(ggplot_build(plot)$data[[1]]$count)*1.1))
}
```

```{r pop1}
# Histogram: original data
plot_histogram(demographics_data, 
               TotalPop, 
               x_label = "Total population",
               title_string = "Histogram: original data")

# Histogram: log10 transformed data
plot_histogram(demographics_data %>% 
                 mutate(TotalPop = log10(TotalPop)),
               TotalPop,
               x_label = "log10(population)",
               title_string = "Log10 transformed population data is better for visualization")
```

```{r normal_pop}
shapiro.test(as_vector(demographics_data %>%
                         select(TotalPop)))

ggplot(demographics_data) +
  geom_qq(aes(sample = TotalPop)) +
  geom_qq_line(aes(sample = TotalPop))

shapiro.test(as_vector(demographics_data %>% 
                         mutate(TotalPop = log10(TotalPop)) %>%
                         select(TotalPop)))

ggplot(demographics_data %>% 
                         mutate(TotalPop = log10(TotalPop)) %>%
                         select(TotalPop)) +
  geom_qq(aes(sample = TotalPop)) +
  geom_qq_line(aes(sample = TotalPop))
```

It's apparent from both the Shapiro-Wilk test (null hypothesis is that the data
is normally distributed, we thus reject the null hypothesis in this case) and
the qqplot that the date is not normally distributed. But the log10 +
transformation clearky make the dataset follow normality more than the original 
data.

See [link](https://ggplot2.tidyverse.org/reference/geom_qq.html) for more on 
QQ-plots.

### Correlation between population and proportion of white ethnicity?

```{r func2, echo = FALSE}
plot_scatterplot <- function(data, column1, column2, y_label, x_label, 
                             title_string = "Scatterplot", 
                             subtitle_string = NULL, caption_string = NULL, 
                             smooth = FALSE) {
  # Library
  library(tidyverse)
  
  # Quote column names
  column1 <- enquo(column1)
  column2 <- enquo(column2)
  
  # Scatter plot
  plot <- data %>%
    ggplot(aes(!!column1, !!column2)) +
    geom_point() +
    xlab(x_label) +
    ylab(y_label) +
    labs(
      title = title_string,
      subtitle = subtitle_string,
      caption = caption_string
    ) +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(color = "black")
    )
  
  # Plots smoothing curve if smooth = TRUE
  if (smooth == TRUE) {
    plot +
      geom_smooth(method = "lm", color = "red", se = FALSE)
  }
  else {
    plot
  }
}
```

Plot `White` against `TotalPop`.

```{r scatter1}
# White vs original TotalPop
plot_scatterplot(demographics_data,
                 column1 = White,
                 column2 = TotalPop,
                 x_label = "White (%)",
                 y_label = "Population")

# White vs log10(TotalPop)
plot_scatterplot(demographics_data %>% 
                   mutate(TotalPop = log10(TotalPop)),
                 column1 = White,
                 column2 = TotalPop,
                 x_label = "White (%)",
                 y_label = "log10(population)")
```

There is no correlation between a county's population and the proportion of
white people living in the county.
